---
title: "Using the Stan Math C++ Library"
author: "Stan Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the Stan Math C++ Library}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: inline
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
Sys.setenv(USE_CXX14 = "1")
set.seed(12345)
```

# Using the **StanHeaders** Package from Other R Packages

The **StanHeaders** package contains no R functions. To use the Stan Math Library
in other packages, it is often sufficient to specify
```
LinkingTo: StanHeaders (>= 2.19.0)
```
in the DESCRIPTION file of another package and
```
CXX_STD = CXX14
```
in the src/Makevars and src/Makevars.win files. If, in addition, the other package needs to utilize 
the MCMC, optimization, variational inference, or parsing facilities of the Stan Library, then it is 
also necessary to include the `src` directory of **StanHeaders** in the other package's `PKG_CPPFLAGS`
in the src/Makevars and src/Makevars.win files with something like
```
STANHEADERS_SRC = $(shell "$(R_HOME)/bin$(R_ARCH_BIN)/Rscript" -e "message()" \
  -e "cat(system.file('include', 'src', package = 'StanHeaders', mustWork = TRUE))" \
  -e "message()" | grep "StanHeaders")
PKG_CPPFLAGS = -I"$(STANHEADERS_SRC)"
```
and put `GNU make` in the `SystemRequirements:` field of the package's DESCRIPTION file. Finally,
if integrating systems of stiff Ordinary Differential Equations, then you need to link to the
shared library of StanHeaders as is shown below.

# Using Higher-Order Functions in the **StanHeaders** Package

This section will demonstrate how to use some of the C++ functions in the **StanHeaders** package.

## Derivatives and Minimization

The following is a toy example of using the Stan Math library via `Rcpp::sourceCpp`:
to minimize the function 
$$\left(\mathbf{x} - \mathbf{a}\right)^\top \left(\mathbf{x} - \mathbf{a}\right)$$
which has a global minimum when $\mathbf{x} = \mathbf{a}$. To find this minimum with autodifferentiation, 
we need to define the objective function. Then, its gradient with respect to $\mathbf{x}$, which we know is 
$2\left(\mathbf{x} - \mathbf{a}\right)$ in this case, can be calculated by autodifferentiation. At the 
optimum (or on the way to the optimum), we might want to evaluate the Hessian matrix, which we know is 
$2\mathbf{I}$, but would need an additional function to evaluate it via autodifferentiation. Finally, one 
could reconceptualize the problem as solving a homogeneous system of equations where the gradient is set 
equal to a vector of zeros. The `stan::math::algebra_solver` function can solve such a system using 
autodifferentiation to obtain the Jacobian, which we know to be the identity matrix in this case.

Here is C++ code that does all of the above, except for the part of finding the optimum, which is done
using the R function `optim` below.
```{Rcpp}
// [[Rcpp::depends(BH)]]
// [[Rcpp::depends(RcppEigen)]]
// [[Rcpp::depends(StanHeaders)]]
#include <stan/math/fwd/mat/fun/dot_self.hpp>    // stuff from fwd/ must come first
#include <stan/math/mix/mat/functor/hessian.hpp> // then stuff from mix/ must come next
#include <stan/math.hpp>                         // finally pull in everything from rev/ & prim/
#include <Rcpp.h>
#include <RcppEigen.h>

// [[Rcpp::plugins(cpp14)]]

/* Objective function */

// [[Rcpp::export]]
auto f(Eigen::VectorXd x, Eigen::VectorXd a) { // objective function in doubles
  using stan::math::dot_self;                  // dot_self() is a dot product with self
  return dot_self( (x - a).eval() );           // .eval() yields a Eigen::VectorXd
}

/* Gradient */

// [[Rcpp::export]]
auto g(Eigen::VectorXd x, Eigen::VectorXd a) {  // gradient by AD using Stan
  double fx;
  Eigen::VectorXd grad_fx;
  using stan::math::dot_self;
  stan::math::gradient([&a](auto x) { return dot_self( (x - a).eval() ); },
                       x, fx, grad_fx);
  return grad_fx;
}

/* Hessian */

// [[Rcpp::export]]
auto H(Eigen::VectorXd x, Eigen::VectorXd a) { // Hessian by AD using Stan
  double fx;
  Eigen::VectorXd grad_fx;
  Eigen::MatrixXd H;
  using stan::math::dot_self;
  using stan::math::subtract; // necessary to get the type promotion correct
  stan::math::hessian([&a](auto x) { return dot_self(subtract(x, a)); },
                      x, fx, grad_fx, H);
  return H;
}

/* Jacobian */

// [[Rcpp::export]]
auto J(Eigen::VectorXd x, Eigen::VectorXd a) { // not actually used
  Eigen::VectorXd fx;
  Eigen::MatrixXd J;
  using stan::math::dot_self;
  stan::math::jacobian([&a](auto x) {
    return (2 * (x - a)).eval();
  }, x, fx, J);
  return J;
}

struct equations_functor {
  template <typename T0, typename T1>
  inline Eigen::Matrix<T0, Eigen::Dynamic, 1>
  operator()(const Eigen::Matrix<T0, Eigen::Dynamic, 1>& x,
             const Eigen::Matrix<T1, Eigen::Dynamic, 1>& theta,
             const std::vector<double>& x_r, const std::vector<int>& x_i,
             std::ostream* pstream__) const {
    return 2 * (x - stan::math::to_vector(x_r)).eval();
  }
};

// [[Rcpp::export]]
auto solution(Eigen::VectorXd a, Eigen::VectorXd guess) {
  Eigen::VectorXd theta;
  auto x_r = stan::math::to_array_1d(a);
  equations_functor f;
  auto x = stan::math::algebra_solver(f, guess, theta, x_r, {});
  return x;
}
```

In this compiled RMarkdown document, the **knitr** package has exported functions `f`, `g`, `H`, `J` 
and `solution` (but not `equations_functor`) to R's global environment using the `sourceCpp` function 
in the **Rcpp** package, so that they can now be called from R. Here we find the optimum starting from 
a random point in three dimensions:
```{r}
x <- optim(rnorm(3), fn = f, gr = g, a = 1:3, method = "BFGS", hessian = TRUE)
x$par
x$hessian
H(x$par, a = 1:3)
J(x$par, a = 1:3)
solution(a = 1:3, guess = rnorm(3))
```

# Integrals and Ordinary Differential Equations

The Stan Math library can do one-dimensional numerical integration and can solve stiff
and non-stiff systems of differential equations, such as the harmonic oscillator example
below. Solving stiff systems utilizes the CVODES package, which is included in **StanHeaders**, 
but requires linking against its shared library with the following R code.

```{r}
StanHeaders_pkg_libs <- system.file(ifelse(.Platform$OS.type == "windows", "libs", "lib"), 
                                    .Platform$r_arch, package = "StanHeaders")
Sys.setenv(PKG_LIBS = paste(paste0("-L", shQuote(StanHeaders_pkg_libs)), "-lStanHeaders"))
```

```{Rcpp}
// [[Rcpp::depends(BH)]]
// [[Rcpp::depends(RcppEigen)]]
// [[Rcpp::depends(StanHeaders)]]
#include <stan/math.hpp>                         // pulls in everything from rev/ and prim/
#include <Rcpp.h>
#include <RcppEigen.h>

// [[Rcpp::plugins(cpp14)]]

/* Definite integrals */

// [[Rcpp::export]]
double Cauchy(double scale) {
  std::vector<double> theta;
  auto half = stan::math::integrate_1d([](auto x, auto xc, auto theta,
                                          auto x_r, auto x_i, auto msgs) {
    return exp(stan::math::cauchy_lpdf(x, 0, x_r[0]));
  }, -scale, scale, theta, {scale}, {}, Rcpp::Rcout, 1e-7);
  return half * 2; // should equal 1 for any positive scale
}

/* Ordinary Differential Equations */

// [[Rcpp::export]]
auto nonstiff(Eigen::MatrixXd A, Eigen::VectorXd y0) {
  using stan::math::integrate_ode_rk45;
  using stan::math::to_vector;
  using stan::math::to_array_1d;
  std::vector<double> theta;
  auto y = integrate_ode_rk45([&A](auto t, auto y, 
                                   auto theta, auto x_r, auto x_i, std::ostream *msgs) {
    return to_array_1d( (A * to_vector(y)).eval() );
  }, to_array_1d(y0), 0, {1, 2}, theta, {}, {});
  return to_vector(y[0]) - stan::math::matrix_exp(A) * y0; // should be numerically zero
}

// [[Rcpp::export]]
auto stiff(Eigen::MatrixXd A, Eigen::VectorXd y0) { // not actually stiff
  using stan::math::integrate_ode_bdf;              // but use the stiff solver anyways
  using stan::math::to_vector;
  using stan::math::to_array_1d;
  std::vector<double> theta;
  auto y = integrate_ode_bdf([&A](auto t, auto y, 
                                  auto theta, auto x_r, auto x_i, std::ostream *msgs) {
    return to_array_1d( (A * to_vector(y)).eval() );
  }, to_array_1d(y0), 0, {1, 2}, theta, {}, {});
  return to_vector(y[0]) - stan::math::matrix_exp(A) * y0; // should be numerically zero
}
```

Again, in this compiled RMarkdown document, the **knitr** package has exported the
`Cauchy`, `nonstiff` and `stiff` functions to R's global environment using the
`sourceCpp` function in the **Rcpp** package so that they can be called from R.

First, we numerically integrate the Cauchy PDF over its interquartile range --- which has
an area of $\frac{1}{2}$ --- that we then double to verify that it is almost within machine
precision of $1$.

```{r}
all.equal(1, Cauchy(rexp(1)), tol = 1e-15)
```

Next, we consider the system of differential equations
$$\frac{d}{dt}\mathbf{y} = \mathbf{A}\mathbf{y}$$
where $\mathbf{A}$ is a square matrix such as that for a simple harmonic oscillator

$$\mathbf{A} = \begin{bmatrix}0 & 1 \\ -1 & -\theta\end{bmatrix}$$
for $\theta \in \left(0,1\right)$. The solution for $\mathbf{y}_t = e^{t\mathbf{A}}\mathbf{y}_0$ can 
be obtained via the matrix exponential function, which is available in the Stan Math Library, but it
can also be obtained numerically using a fourth-order Runge-Kutta solver, which is 
appropriate for non-stiff systems of ODEs, such as this one. However, it is possible,
albeit less efficient in this case, to use the backward-differentiation formula solver
for stiff systems of ODEs. In both cases, we calculate the difference between the
analytical solution and the numerical one, and the stiff version does produce somewhat
better accuracy in this case.
```{r}
A <- matrix(c(0, -1, 1, -runif(1)), nrow = 2, ncol = 2)
y0 <- rexp(2)
all.equal(nonstiff(A, y0), c(0, 0), tol = 1e-5)
all.equal(   stiff(A, y0), c(0, 0), tol = 1e-8)
```

# Map and Parellelization

The Stan Math Library includes the `map_rect` function, which applies a function to each element
of rectangular arrays and returns a vector, making it a bit like a restricted version of R's `sapply` 
function. However, `map_rect` can also be executed in parallel by defining the pre-processor
directive `STAN_THREADS` and then setting the `STAN_NUM_THREADS` environmental variable to be the 
number of threads to use, as in
```{r}
Sys.setenv(PKG_CXXFLAGS = "-DSTAN_THREADS")
Sys.setenv(STAN_NUM_THREADS = 2) # specify -1 to use all available cores
```

Below is C++ code to test whether an integer is prime, using a rather brute-force algorithm
and running it in parallel via `map_rect`.
```{Rcpp}
// [[Rcpp::depends(BH)]]
// [[Rcpp::depends(RcppEigen)]]
// [[Rcpp::depends(StanHeaders)]]
#include <stan/math.hpp>                         // pulls in everything from rev/ and prim/
#include <Rcpp.h>
#include <RcppEigen.h>

// [[Rcpp::plugins(cpp14)]]

// see https://en.wikipedia.org/wiki/Primality_test#Pseudocode
struct is_prime {
  is_prime() {}
  template <typename T1, typename T2>
  auto
  operator()(const Eigen::Matrix<T1, Eigen::Dynamic, 1>& eta,
             const Eigen::Matrix<T2, Eigen::Dynamic, 1>& theta,
             const std::vector<double>& x_r, const std::vector<int>& x_i,
             std::ostream* msgs = 0) const {
    Eigen::VectorXd res(1); // can only return double or var vectors
    int n = x_i[0];
    if (n <= 3) {
      res.coeffRef(0) = n > 1;
      return res;
    } else if ( (n % 2 == 0) || (n % 3 == 0) ) {
      res.coeffRef(0) = false;
      return res;
    }
    int i = 5;
    while (i * i <= n) {
      if ( (n % i == 0) || (n % (i + 2) == 0) ) {
        res.coeffRef(0) = false;
        return res;
      }
      i += 6;
    }
    res.coeffRef(0) = true;
    return res;
  }
};

/* parallelization */
// [[Rcpp::export]]
auto psapply(std::vector<std::vector<int> > n) {
  std::vector<Eigen::VectorXd> eta(n.size()); // these all have to be the same size
  Eigen::VectorXd theta;
  std::vector<std::vector<double> > x_d(n.size());
  return stan::math::map_rect<0, is_prime>(theta, eta, x_d, n, &Rcpp::Rcout);
}
```

Since the signature for `n` is a `std::vector<std::vector<int> >`, we have to pass
it from R as a list (which is converted to the outer `std::vector<>`) of integer
vectors (which is converted to the inner `std::vector<int>`) that happen to be of
size one in this case.
```{r}
odd <- seq.int(from = 2^25 - 1, to = 2^26 - 1, by = 2)
tail(psapply(n = as.list(odd))) == 1 # check your process manager while this is running
```
Thus, $2^{26} - 5 = 67,108,859$ is a prime number.

# Defining a Stan Model in C++

The Stan _language_ does not have much support for sparse matrices for a variety of
reasons. Essentially the only applicable function is `csr_matrix_times_vector`, which 
pre-multiplies a vector by a sparse matrix in compressed row storage by taking as arguments 
its number of rows, columns, non-zero values, column indices of non-zero values, and
locations where the non-zero values start in each row. While the 
`csr_matrix_times_vector` function could be used to implement the example below,
we illustrate how to use the sparse data structures in the **Matrix** and **RcppEigen**
packages in a Stan model written in C++, which could easily be extended to more complicated
models with sparse data structures.

Our C++ file for the log-likelihood of a linear model with a sparse design matrix reads as
```{r, echo = FALSE, comment = ""}
cat(readLines("sparselm_stan.hpp"), sep = "\n")
```

To use it from R, we call the `exposeClass` function in the **Rcpp** package with
the necessary arguments and then call `sourceCpp` on the file it wrote in the
temporary directory:
```{r, message = FALSE}
library(Rcpp)
tf <- tempfile(fileext = "Module.cpp")
exposeClass("sparselm_stan",
      constructors = list(c("Eigen::Map<Eigen::SparseMatrix<double> >", 
                            "Eigen::VectorXd")),
      fields = c("X", "y"),
      methods = c("log_prob<>", "gradient<>"),
      rename = c(log_prob = "log_prob<>", gradient = "gradient<>"),
      header = c("// [[Rcpp::depends(BH)]]",
                 "// [[Rcpp::depends(RcppEigen)]]",
                 "// [[Rcpp::depends(StanHeaders)]]",
                 "// [[Rcpp::plugins(cpp14)]]",
                 paste0("#include <", file.path(getwd(), "sparselm_stan.hpp"), ">")),
      file = tf,
      Rfile = FALSE)
Sys.unsetenv("PKG_CXXFLAGS") # don't specify -DSTAN_THREADS if you are not using them
Sys.setenv(PKG_CPPFLAGS = paste0("-I",
                                 system.file("include", "src", 
                                             package = "StanHeaders", mustWork = TRUE)))
sourceCpp(tf)
sparselm_stan
```
At this point, we need a sparse design matrix and (dense) outcome vector to pass to
the constructor. The former can be created with a variety of functions in the **Matrix**
package, such as
```{r}
dd <- data.frame(a = gl(3, 4), b = gl(4, 1, 12))
X <- Matrix::sparse.model.matrix(~ a + b, data = dd)
X
```
Finally, we call the `new` function in the **methods** package, which essentially
calls our C++ constructor and provides an R interface to the instantiated object,
which contains the `log_prob` and `gradient` methods we defined and can be called
with arbitrary inputs.
```{r}
sm <- new(sparselm_stan, X = X, y = rnorm(nrow(X)))
sm$log_prob(c(beta = rnorm(ncol(X)), log_sigma = log(pi)))
round(sm$gradient(c(beta = rnorm(ncol(X)), log_sigma = log(pi))), digits = 4)
```

