---
title: "Approximate leave-future-out cross-validation for time series models"
author: "Paul Bürkner, Aki Vehtari, Jonah Gabry"
date: "`r Sys.Date()`"
output:
  html_vignette:
    toc: yes
encoding: "UTF-8"
params:
  EVAL: !r identical(Sys.getenv("NOT_CRAN"), "true")
---

<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Approximate leave-future-out cross-validation for time series models}
-->

```{r settings, child="children/SETTINGS-knitr.txt"}
```

```{r more-knitr-ops, include=FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  message = FALSE, 
  warning = FALSE
)
```

## Introduction

One of the most common goals of a time series analysis is to use the observed
series to inform predictions for future observations. We will refer to this task
of predicting a sequence of $M$ future observations as $M$-step-ahead prediction
($M$-SAP). Fortunately, once we have fit a model and can sample from the
posterior predictive distribution, it is straightforward to generate predictions
as far into the future as we want. It is also straightforward to evaluate the
$M$-SAP performance of a time series model by comparing the predictions to the
observed sequence of $M$ future data points once they become available.

Unfortunately, we are often in the position of having to use a model to inform
decisions _before_ we can collect the future observations required for assessing
the predictive performance. If we have many competing models we may also need to
first decide which of the models (or which combination of the models) we should
rely on for predictions. In these situations the best we can do is to use
methods for approximating the expected predictive performance of our models
using only the observations of the time series we already have.

If there were no time dependence in the data or if the focus is to assess
the non-time-dependent part of the model, we could use methods like
leave-one-out cross-validation (LOO-CV). For a data set with $N$ observations,
we refit the model $N$ times, each time leaving out one of the $N$ observations
and assessing how well the model predicts the left-out observation. LOO-CV is
very expensive computationally in most realistic settings, but the Pareto
smoothed importance sampling (PSIS, Vehtari et al, 2017ab) algorithm provided by
the *loo* package allows for approximating exact LOO-CV with PSIS-LOO-CV.
PSIS-LOO-CV requires only a single fit of the full model and comes with
diagnostics for assessing the validity of the approximation.

With a time series we can do something similar to LOO-CV but, except in a few
cases, it does not make sense to leave out observations one at a time because
then we are allowing information from the future to influence predictions of the
past (i.e., times $t + 1, t+2, \ldots$ should not be used to predict for time
$t$). To apply the idea of cross-validation to the $M$-SAP case, instead of
leave-*one*-out cross-validation we need some form of leave-*future*-out
cross-validation (LFO-CV). As we will demonstrate in this case study, LFO-CV
does not refer to one particular prediction task but rather to various possible
cross-validation approaches that all involve some form of prediction for new
time series data. Like exact LOO-CV, exact LFO-CV requires refitting the model
many times to different subsets of the data, which is computationally very
costly for most
nontrivial examples, in particular for Bayesian analyses where refitting the
model means estimating a new posterior distribution rather than a point
estimate.

Although PSIS-LOO-CV provides an efficient approximation to exact LOO-CV, until
now there has not been an analogous approximation to exact LFO-CV that
drastically reduces the computational burden while also providing informative
diagnostics about the quality of the approximation. In this case study we
present PSIS-LFO-CV, an algorithm that typically only requires refitting the
time-series model a small number times and will make LFO-CV tractable for many
more realistic applications than previously possible. 

More details can be found in our paper about approximate LFO-CV
(Bürkner, Gabry, & Vehtari, 2019), which is available as a preprint on arXiv (https://arxiv.org/abs/1902.06281).


## $M$-step-ahead predictions

Assume we have a time series of observations 
$y = (y_1, y_2, \ldots, y_N)$ 
and let $L$ be the _minimum_ number of observations from the series that
we will require before making predictions for future data. Depending on the
application and how informative the data is, it may not be possible to make
reasonable predictions for $y_{i}$ based on $(y_1, \dots, y_{i-1})$ until $i$ is
large enough so that we can learn enough about the time series to predict future
observations. Setting $L=10$, for example, means that we will only assess
predictive performance starting with observation $y_{11}$, so that we
always have at least 10 previous observations to condition on.

In order to assess $M$-SAP performance we would like to compute the 
predictive densities

$$
p(y_{i+1:M} \,|\, y_{1:i}) = p(y_i, \ldots, y_{i + M - 1} \,|\, y_{1},...,y_{i-1}) 
$$ 

for each $i \in \{L + 1, \ldots, N - M + 1\}$. The quantities 
$p(y_{i+1:M} \,|\, y_{1:i})$ can be computed with the help of the posterior
distribution $p(\theta \,|\, y_{1:i})$ of the parameters $\theta$ conditional on
only the first $i-1$ observations of the time-series:

$$
p(y_{i+1:M} \,| \, y_{1:i}) = 
  \int p(y_{i+1:M} \,| \, y_{1:i}, \theta) \, p(\theta\,|\,y_{1:i}) \,d\theta. 
$$

Having obtained $S$ draws $(\theta_{1:i}^{(1)}, \ldots, \theta_{1:i}^{(S)})$ 
from the posterior distribution $p(\theta\,|\,y_{1:i})$, we can estimate 
$p(y_{i+1:M} | y_{1:i})$ as

$$
p(y_{i+1:M} \,|\, y_{1:i}) \approx \frac{1}{S}\sum_{s=1}^S p(y_{i+1:M} \,|\, y_{1:i}, \theta_{1:i}^{(s)}).
$$

In the following, we consider factorizable models in which the response values
are conditionally independent given the parameters and the likelihood can be
written in the familiar form

$$
p(y \,|\, \theta) = \prod_{i=1}^N p(y_i \,|\, \theta).
$$

In this case, $p(y_{i+1:M} \,|\, y_{1:i}, \theta)$ reduces to 
$$
p(y_{i+1:M} \,|\, \theta) = \prod_{j = i}^{i + M -1} p(y_j \,|\, \theta),
$$
due to the assumption of conditional independence between $y_{i+1:M}$ and
$y_{1:i}$ given $\theta$.

Non-factorizable models, which do not make this assumption, are discussed in a
separate vignette [_Leave-one-out cross-validation for non-factorizable models_](http://mc-stan.org/loo/articles/loo2-non-factorizable.html).

## Approximate $M$-SAP using importance-sampling {#approximate_MSAP}

Unfortunately, the math above makes use of the posterior distributions from many
different fits of the model to different subsets of the data. That is, to obtain
the predictive density $p(y_{i+1:M} \,|\, y_{1:i})$ requires fitting a model to
only the first $i-1$ data points, and we will need to do this for every value of
$i$ under consideration (all $i \in \{L + 1, \ldots, N - M + 1\}$).

To reduce the number of models that need to be fit for the purpose of obtaining
each of the densities $p(y_{i+1:M} \,|\, y_{1:i})$, we propose the following
algorithm. Starting with $i = N - M + 1$, we approximate each 
$p(y_{i+1:M} \,|\, y_{1:i})$ using Pareto smoothed importance sampling (PSIS,
Vehtari et al, 2017ab):

$$
 p(y_{i+1:M} \,|\, y_{1:i}) \approx
   \frac{ \sum_{s=1}^S w_i^{(s)}\, p(y_{i+1:M} \,|\, \theta^{(s)})}{ \sum_{s=1}^S w_i^{(s)}},
$$

where $w_i^{(s)}$ are importance weights and $\theta^{(s)}$ are draws from the
posterior distribution based on _all_ observations. To obtain $w_i^{(s)}$, we
first compute the raw importance ratios

$$
r_i^{(s)} \propto \frac{1}{\prod_{j = i}^N p(y_j \,|\, \,\theta^{(s)})},
$$

and then stabilize them using PSIS.

We then _decrease_ $i$ by $1$ (i.e., we move backwards in time) and repeat the
process. At some observation $i$, the variability of importance ratios
$r_i^{(s)}$ will become too large and importance sampling fails. We will refer
to this particular value of $i$ as $i^\star_1$. To identify the value of
$i^\star_1$ we check for which value of $i$ does the estimated shape parameter
$k$ of the generalized Pareto distribution first crosses a certain threshold.
Only then do we refit the model using only observations before $i^\star_1$ and
then restart the process. In some cases we may only need to refit once and in
other cases we will find a value $i^\star_2$ that requires a second refitting,
maybe an $i^\star_3$ that requires a third refitting, and so on. We repeat the
refitting as few times as is required (only if $k > \text{threshold}$) until we
arrive at $i = L+1$. Recall that $L$ is the minimum number of observations we
have deemed acceptable for making predictions (setting $L=0$ means predictions
of all observations should be computed).

For LOO, we recommend to use a threshold of $0.7$ (Vehtari et al, 2017ab), but
for LFO, the threshold should be smaller as influential observations will affect
all pointwise LFO approximations until the next refit and thus much stronger
than in LOO. For the present case study, we choose a threshold of $0.6$, 
which also seems to be generally reasonable threshold for LFO (Bürkner et al. 2019).


## Autoregressive models

Autoregressive (AR) models are some of the most commonly used time-series models. 
An AR(p) model ---an autoregressive model of order $p$--- can be defined as

$$
y_i = \eta_i + \sum_{k = 1}^p \varphi_k y_{i - k} + \varepsilon_i,
$$

where $\eta_i$ is the linear predictor for the $i$th observation, $\phi_k$ are
the autoregressive parameters and $\varepsilon_i$ are pairwise independent
errors, which are usually assumed to be normally distributed with equal variance
$\sigma^2$. The model implies a recursive formula that allows for computing the
right-hand side of the above equation for observation $i$ based on the values of
the equations for previous observations.

## Case Study: Annual measurements of the level of Lake Huron

To illustrate the application of PSIS-LFO-CV for estimating expected $M$-SAP
performance, we will fit a model for 98 annual measurements of the water level
(in feet) of [Lake Huron](https://en.wikipedia.org/wiki/Lake_Huron) from the
years 1875--1972. This data set is found in the **datasets** R package, which is
installed automatically with **R**. 

In addition to the **loo** package, for this analysis we will use the **brms**
interface to Stan to generate a Stan program and fit the model, and also the
**bayesplot** and **ggplot2** packages for plotting.

```{r pkgs, cache=FALSE}
library("loo")
library("brms")
library("bayesplot")
library("ggplot2")
color_scheme_set("brightblue")
theme_set(theme_default())

CHAINS <- 4
SEED <- 1234
set.seed(SEED)
```

Before fitting a model, we will first put the data into a data frame and then
look at the time series.

```{r hurondata}
N <- length(LakeHuron)
df <- data.frame(
  y = as.numeric(LakeHuron),
  year = as.numeric(time(LakeHuron)),
  time = 1:N
)

# save plot labels to reuse them
plot_labs <- labs(
  y = "Water Level (ft)", 
  x = "Year",
  title = "Water Level in Lake Huron (1875-1972)"
) 

ggplot(df, aes(x = year, y = y)) + 
  geom_point(size = 1) +
  plot_labs
```

The above plot shows rather strong autocorrelation of the time-series as well as
some trend towards lower levels for later points in time. 

We can specify an AR(4) model for these data using the **brms** package as
follows:

```{r fit, results = "hide"}
control <- list(adapt_delta = 0.95)
fit <- brm(
  y ~ 1, 
  data = df, 
  autocor = cor_ar(~time, p = 4), 
  prior = prior(normal(0, 0.5), class = "ar"),
  control = control, 
  seed = SEED, 
  chains = CHAINS
)
```

The model implied predictions along with the observed values can be plotted,
which reveals a rather good fit to the data.

```{r plotpreds, cache = FALSE}
preds <- posterior_predict(fit)
preds <- cbind(
  Estimate = colMeans(preds), 
  Q5 = apply(preds, 2, quantile, probs = 0.05),
  Q95 = apply(preds, 2, quantile, probs = 0.95)
)

ggplot(cbind(df, preds), aes(x = year, y = Estimate)) +
  geom_smooth(aes(ymin = Q5, ymax = Q95), stat = "identity", size = 0.5) +
  geom_point(aes(y = y)) + 
  labs(subtitle = "Mean (blue) and 90% predictive intervals (gray) vs. observed data (black)") +
  plot_labs
```

To allow for reasonable predictions of future values, we will require at least
$L = 20$ historical observations (20 years) to make predictions.

```{r setL}
L <- 20
```

We first perform approximate leave-one-out cross-validation (LOO-CV) for the
purpose of later comparison with exact and approximate LFO-CV for the 1-SAP
case.

```{r loo1sap, cache = FALSE}
loo_cv <- loo(log_lik(fit)[, (L + 1):N])
print(loo_cv)
```

## 1-step-ahead predictions leaving out all future values

The most basic version of $M$-SAP is 1-SAP, in which we predict only one step
ahead. In this case, $y_{i+1:M}$ simplifies to $y_{i}$ and the LFO-CV algorithm
becomes considerably simpler than for larger values of $M$.

### Exact 1-step-ahead predictions

Before we compute approximate LFO-CV using PSIS we will first compute 
exact LFO-CV for the 1-SAP case so we can use it as a benchmark later. 
The initial step for the exact computation is to calculate the log-predictive 
densities by refitting the model many times:

```{r exact_loglik, results="hide"}
loglik_exact <- matrix(nrow = nsamples(fit), ncol = N)
for (i in N:max(L + 1, 2)) {
  fit_i <- update(fit, newdata = df[-(i:N), ], recompile = FALSE)
  loglik_exact[, i] <- log_lik(fit_i, newdata = df[1:i, ])[, i]
}
```

Then we compute the exact expected log predictive density (ELPD):

```{r helpers}
# some helper functions we'll use throughout

# more stable than log(sum(exp(x))) 
log_sum_exp <- function(x) {
  max_x <- max(x)  
  max_x + log(sum(exp(x - max_x)))
}

# more stable than log(mean(exp(x)))
log_mean_exp <- function(x) {
  log_sum_exp(x) - log(length(x))
}

# compute log of raw importance ratios
# sums over observations *not* over posterior samples
sum_log_ratios <- function(ll, ids = NULL) {
  if (!is.null(ids)) ll <- ll[, ids, drop = FALSE]
  - rowSums(ll)
}

# for printing comparisons later
rbind_print <- function(...) {
  round(rbind(...), digits = 2)
}
```


```{r exact1sap, cache = FALSE}
exact_elpds_1sap <- apply(loglik_exact, 2, log_mean_exp)
exact_elpd_1sap <- c(ELPD = sum(exact_elpds_1sap[-(1:L)]))

rbind_print(
  "LOO" = loo_cv$estimates["elpd_loo", "Estimate"],
  "LFO" = exact_elpd_1sap
)
```

We see that the ELPD from LFO-CV for 1-step-ahead predictions is lower than the
ELPD estimate from LOO-CV, which should be expected since LOO-CV is making use
of more of the time series. That is, since the LFO-CV approach only uses
observations from before the left-out data point but LOO-CV uses _all_ data
points other than the left-out observation, we should expect to see the larger
ELPD from LOO-CV.

### Approximate 1-step-ahead predictions

For illustrative purposes, we will first compute approximate 1-SAP without any
refitting, even when the Pareto $k$ estimate is too large. This will of course
be a poor approximation to exact 1-SAP, in particular if the Pareto $k$
estimates increase rather quickly beyond the threshold up to which PSIS tends to
produce stable results.

```{r loglik}
approx_elpds_1sap_no_refit <- rep(NA, N)
loglik <- log_lik(fit)
ids <- N:(L + 1)
ks <- NULL

for (i in ids) {
  logratio <- sum_log_ratios(loglik, i:N)
  psis_part <- suppressWarnings(psis(logratio))
  ks <- c(ks, pareto_k_values(psis_part))
  lw_i <- weights(psis_part, normalize = TRUE)[, 1]
  approx_elpds_1sap_no_refit[i] <- log_sum_exp(lw_i + loglik[, i])
}
```

If we plot the Pareto $k$ estimates we can see that they do start to increase 
quickly and that once we work backwards to about halfway through the data points 
we get $k > 0.6$. This is indicated by the vertical line in the plot below, which 
corresponds to the point $i^\star_1$ discussed in Section
[Approximate $M$-SAP using importance-sampling](#approximate_MSAP) above.

```{r plot_ks}
plot_ks <- function(ks, ids, thres = 0.6) {
  dat_ks <- data.frame(ks = ks, ids = ids)
  ggplot(dat_ks, aes(x = ids, y = ks)) + 
    geom_point(aes(color = ks > thres), shape = 3, show.legend = FALSE) + 
    geom_hline(yintercept = thres, linetype = 2, color = "red2") + 
    scale_color_manual(values = c("cornflowerblue", "darkblue")) + 
    labs(x = "Data point", y = "Pareto k") + 
    ylim(-0.5, 1.5)
}
```

```{r plotk1sap, cache = FALSE}
plot_ks(ks, ids)
```

Consequently, it is not surprising that the resulting approximate ELPD for 1-SAP 
is far away from the exact ELPD we computed above:

```{r compare1sap, cache = FALSE}
approx_elpd_1sap_no_refit <- 
  c(ELPD = sum(approx_elpds_1sap_no_refit, na.rm = TRUE))
rbind_print(
  "approx LFO (no refit)" = approx_elpd_1sap_no_refit,
  "exact LFO" = exact_elpd_1sap
)
```

Next, we compute approximate 1-SAP with refit at observations where 
the Pareto $k$ estimate exceeds the threshold of $0.6$.

```{r setkthresh}
k_thres <- 0.6
```

The code becomes a little bit more involved to handle the refitting procedure.
Note that we can compute exact 1-SAP at the refitting points, which comes with
no additional computational costs since we had to refit the model anyway.

```{r refit_loglik, results="hide"}
loglik <- matrix(nrow = nsamples(fit), ncol = N)
approx_elpds_1sap <- rep(NA, N)

fit_part <- fit
ids <- N:(L + 1)
i_refit <- N
refits <- NULL
ks <- NULL

for (i in ids) {
  loglik[, i] <- log_lik(fit_part)[, i]
  logratio <- sum_log_ratios(loglik, i:i_refit)
  psis_part <- suppressWarnings(psis(logratio))
  k <- pareto_k_values(psis_part)
  ks <- c(ks, k)
  if (k > k_thres) {
    # refit the model based on the first i-1 observations
    i_refit <- i - 1
    refits <- c(refits, i)
    fit_part <- update(fit_part, newdata = df[1:(i - 1), ], recompile = FALSE)
    loglik[, i] <- log_lik(fit_part, newdata = df[1:i, ])[, i]
    approx_elpds_1sap[i] <- log_mean_exp(loglik[, i])
  } else {
    lw_i <- weights(psis_part, normalize = TRUE)[, 1]
    approx_elpds_1sap[i] <- log_sum_exp(lw_i + loglik[, i])
  }
} 
```

We see that the final Pareto-$k$-estimates are mostly well below the threshold
and that we only needed to refit the model a few times:

```{r refitsummary1sap, cache=FALSE}
cat("Using threshold ", k_thres, 
    ", model was refit ", length(refits), 
    " times, at observations", refits)

plot_ks(ks, ids)
```

The approximate 1-SAP ELPD is remarkably similar to the exact 1-SAP ELPD
computed above, which indicates our algorithm to compute approximate 1-SAP
worked well for the present data and model.

```{r lfosummary1sap, cache = FALSE}
approx_elpd_1sap <- sum(approx_elpds_1sap, na.rm = TRUE)
rbind_print(
  "approx LFO (no refits)" = approx_elpd_1sap_no_refit,
  "approx LFO (with refits)" = approx_elpd_1sap,
  "exact LFO" = exact_elpd_1sap
)
```

Plotting exact against approximate predictions, we see that no approximation
value deviates far from its exact counterpart, providing further evidence for
the good quality of our approximation.

```{r plot1sap, cache = FALSE}
dat_elpd <- data.frame(
  approx_elpd = approx_elpds_1sap,
  exact_elpd = exact_elpds_1sap
)

ggplot(dat_elpd, aes(x = approx_elpd, y = exact_elpd)) +
  geom_abline(color = "gray30") +
  geom_point(size = 2) +
  labs(x = "Approximate ELPDs", y = "Exact ELPDs")
```

We can also look at the maximum difference and average difference between
the approximate and exact ELPD calculations, which also indicate a ver close approximation:

```{r diffs1sap}
max_diff <- with(dat_elpd, max(abs(approx_elpd - exact_elpd), na.rm = TRUE))
mean_diff <- with(dat_elpd, mean(abs(approx_elpd - exact_elpd), na.rm = TRUE))

rbind_print(
  "Max diff" = round(max_diff, 2), 
  "Mean diff" =  round(mean_diff, 3)
)
```


## $M$-step-ahead predictions leaving out all future values

To illustrate the application of $M$-SAP for $M > 1$, we next compute exact and
approximate LFO-CV for the 4-SAP case.

### Exact $M$-step-ahead predictions

The necessary steps are the same as for 1-SAP with the exception that the
log-density values of interest are now the sums of the log predictive densities
of four consecutive observations. Further, the stability of the PSIS
approximation actually stays the same for all $M$ as it only depends on the
number of observations we leave out, not on the number of observations we
predict.

```{r exact_loglikm, results="hide"}
M <- 4
loglikm <- matrix(nrow = nsamples(fit), ncol = N)
for (i in (N - M + 1):max(L + 1, 2)) {
  fit_i <- update(fit, newdata = df[-(i:N), ], recompile = FALSE)
  ll <- log_lik(fit_i, newdata = df[1:(i + M - 1), ], oos = i:(i + M - 1))
  loglikm[, i] <- rowSums(ll[, i:(i + M - 1)])
}
```

```{r exact4sap, cache = FALSE}
exact_elpds_4sap <- apply(loglikm, 2, log_mean_exp)
(exact_elpd_4sap <- c(ELPD = sum(exact_elpds_4sap, na.rm = TRUE)))
```

### Approximate $M$-step-ahead predictions

Computing the approximate PSIS-LFO-CV for the 4-SAP case is a little bit more
involved than the approximate version for the 1-SAP case, although the
underlying principles remain the same.

```{r refit_loglikm, results="hide"}
loglikm <- loglik <- matrix(nrow = nsamples(fit), ncol = N)
approx_elpds_4sap <- rep(NA, N)
fit_part <- fit
ids <- (N - M + 1):(L + 1)
i_refit <- N - M + 1
refits <- NULL
ks <- NULL

loglik[, (N - M + 2):N] <- log_lik(fit_part)[, (N - M + 2):N]

for (i in ids) {
  oos <- i:(i + M - 1)
  ll <- log_lik(fit_part, newdata = df[1:(i + M - 1), ], oos = oos)
  loglikm[, i] <- rowSums(ll[, i:(i + M - 1)])
  loglik[, i] <- ll[, i]
  logratio <- sum_log_ratios(loglik, i:i_refit)
  psis_part <- suppressWarnings(psis(logratio))
  k <- pareto_k_values(psis_part)
  ks <- c(ks, k)
  if (k > k_thres) {
    # refit the model based on the first i-1 observations
    i_refit <- i - 1
    refits <- c(refits, i)
    fit_part <- update(
      fit_part, newdata = df[1:(i - 1), ], 
      recompile = FALSE
    )
    ll <- log_lik(fit_part, newdata = df[1:(i + M - 1), ], oos = oos)
    loglik[, i] <- ll[, i]
    loglikm[, i] <- rowSums(ll[, i:(i + M - 1)])
    approx_elpds_4sap[i] <- log_mean_exp(loglikm[, i])
  } else {
    lw_i <- weights(psis_part, normalize = TRUE)[, 1]
    approx_elpds_4sap[i] <- log_sum_exp(lw_i + loglikm[, i])
  }
} 
```

Again, we see that the final Pareto-$k$-estimates are mostly well below the
threshold and that we only needed to refit the model a few times: 

```{r refitsummary4sap, cache = FALSE}
cat("Using threshold ", k_thres, 
    ", model was refit ", length(refits), 
    " times, at observations", refits)

plot_ks(ks, ids)
```

The approximate ELPD computed for the 4-SAP case is not as close to its exact
counterpart as in the 1-SAP case.

```{r lfosummary4sap, cache = FALSE}
approx_elpd_4sap <- sum(approx_elpds_4sap, na.rm = TRUE)
rbind_print(
  "Approx LFO" = approx_elpd_4sap,
  "Exact LFO" = exact_elpd_4sap
)
```

Plotting exact against approximate pointwise predictions confirms that, for a 
few specific data points, the approximate predictions underestimate the exact
predictions.

```{r plot4sap, cache = FALSE}
dat_elpd_4sap <- data.frame(
  approx_elpd = approx_elpds_4sap,
  exact_elpd = exact_elpds_4sap
)

ggplot(dat_elpd_4sap, aes(x = approx_elpd, y = exact_elpd)) +
  geom_abline(color = "gray30") +
  geom_point(size = 2) +
  labs(x = "Approximate ELPDs", y = "Exact ELPDs")
```

## Conclusion

In this case study we have shown how to do carry out exact and approximate
leave-future-out cross-validation for $M$-step-ahead prediction tasks. For the
data and model used in our example, the PSIS-LFO-CV algorithm provides reasonably
stable and accurate results despite not requiring us to refit the model nearly as
many times. For more details on approximate LFO-CV, we refer to Bürkner et al. (2019).

<br />

## References

Bürkner P. C., Gabry J., & Vehtari A. (2019). Approximate leave-future-out cross-validation for time series models. [arXiv preprint](https://arxiv.org/abs/1902.06281). 

Vehtari A., Gelman A., & Gabry J. (2017a). Practical Bayesian model evaluation using leave-one-out cross-validation and WAIC. *Statistics and Computing*, 27(5), 1413--1432. \doi:10.1007/s11222-016-9696-4. [Online](http://link.springer.com/article/10.1007/s11222-016-9696-4). [arXiv preprint arXiv:1507.04544](https://arxiv.org/abs/1507.04544).

Vehtari A., Gelman A., & Gabry J. (2017b). Pareto smoothed importance sampling. [arXiv preprint arXiv:1507.02646](https://arxiv.org/abs/1507.02646).

<br />

## Appendix

### Appendix: Session information

```{r sessioninfo}
sessionInfo()
```

### Appendix: Licenses

* Code &copy; 2018, Paul Bürkner, Aki Vehtari, Jonah Gabry (licensed under BSD-3).
* Text &copy; 2018, Paul Bürkner, Aki Vehtari, Jonah Gabry (licensed under CC-BY-NC 4.0).
