---
title: "Interfacing with External C++ Code"
author: "Stan Development Team"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Interfacing with External C++ Code}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(rstan)
knitr::opts_chunk$set(
  echo = TRUE, eval = FALSE, error = TRUE
)
```

Starting with the 2.13 release, it is much easier to use external C++ code in a
Stan program. This vignette briefly illustrates how to do so.

Suppose that you have (part of) a Stan program that involves Fibonacci numbers,
such as
```{stan eval = FALSE, output.var = "fib"}
functions {
  int fib(int n);
  int fib(int n) {
    if (n <= 0) reject("n must be positive");
    return n <= 2 ? 1 : fib(n - 1) + fib(n - 2);
  }
}
model {} // use the fib() function somehow
```
On the second line, we have _declared_ the `fib` function before it is _defined_
in order to call it recursively.

For functions that are not recursive, it is not necessary to declare them before
defining them but it may be advantageous. For example, I often like to hide the
definitions of complicated utility functions that are just a distraction using
the `#include "file"` mechanism
```{stan eval = FALSE, output.var="includes"}
functions {
  real complicated(real a, real b, real c, real d, real e, real f, real g);
#include "complicated.stan"
}
model {} // use the complicated() function somehow
```
This Stan program would have to be parsed using the `stanc_builder` function in
the __rstan__ package rather than the default `stanc` function (which is called
by `sampling` and `stan` internally).

Returning to the Fibonacci example, it is not necessary to define the `fib`
function using the Stan language because Stan programs with functions that are
_declared_ but not _defined_ can use the standard capabilities of the C++
toolchain to provide the function definitions in C++. For example, this program
produces a parser error by default
```{r, eval = TRUE}
mc <-
'
functions { int fib(int n); }
model {} // use the fib() function somehow
'
try(stan_model(model_code = mc, model_name = "parser_error"), silent = TRUE)
```
However, if we specify the `allow_undefined` and `includes` arguments to the
`stan_model` function, and define a `fib` function in the named C++ header file,
then it will parse and compile
```{r}
stan_model(model_code = mc, model_name = "external", allow_undefined = TRUE,
           includes = paste0('\n#include "',
                             file.path(getwd(), 'fib.hpp'), '"\n'))
```
Specifying the `includes` argument is a bit awkward because the C++
representation of a Stan program is written and compiled in a temporary
directory. Thus, the `includes` argument must specify a _full_ path to the
fib.hpp file, which in this case is in the working directory. Also, the
path must be enclosed in double-quotes, which is why single quotes are used
in the separate arguments to the `paste0` function so that double-quotes are
interpreted literally. Finally, the `includes` argument should include newline
characters (`"\n"`) at the start and end. It is possible to specify multiple
paths using additional newline characters or include a "meta-header" file
that contains `#include` directives to other C++ header files.

The result of the `includes` argument is inserted into the C++ file directly
at the end of the lines (as opposed to CmdStan where it is inserted directly
_before_ the following lines)
```{Rcpp, eval = FALSE}
#include <stan/model/model_header.hpp>

namespace some_namespace {

using std::istream;
using std::string;
using std::stringstream;
using std::vector;
using stan::io::dump;
using stan::math::lgamma;
using stan::model::prob_grad;
using namespace stan::math;

typedef Eigen::Matrix<double,Eigen::Dynamic,1> vector_d;
typedef Eigen::Matrix<double,1,Eigen::Dynamic> row_vector_d;
typedef Eigen::Matrix<double,Eigen::Dynamic,Eigen::Dynamic> matrix_d;

static int current_statement_begin__;

// various function declarations and / or definitions
#include "/full/path/to/fib.hpp"
```
Thus, the definition of the `fib` function in the fib.hpp file need not be
enclosed in any particular namespace (which is a random string by default. The
"meta-include" stan/model/model_header.hpp file reads as
```{r, echo = FALSE, eval = TRUE, comment = NA}
cat(readLines(system.file("include", "src", "stan", "model", "model_header.hpp",
                          package = "StanHeaders")), sep = "\n")
```
so the definition of the `fib` function in the fib.hpp file could utilize any
function in the Stan Math Library (without having to prefix function calls with
`stan::math::`), some typedefs to classes in the Eigen matrix algebra library,
plus streams, exceptions, etc. without having to worry about the corresponding
header files. Nevertheless, an external C++ file _may_ contain additional
include directives that bring in class definitions, for example.

Now let's examine the fib.hpp file, which contains the C++ lines
```{Rcpp, eval = FALSE}
int fib(const int&n, std::ostream* pstream__) {
  if (n <= 0) {
    stringstream errmsg;
    errmsg << "n must be positive";
    throw std::domain_error(errmsg.str());
  }
  return n <= 1 ? 1 : fib(n - 1, 0) + fib(n - 2, 0);
}
```
This C++ function is essentially what the preceding user-defined function in the
Stan language
```{stan eval = FALSE, output.var="def"}
int fib(int n) {
  if (n <= 0) reject("n must be positive");
  return n <= 2 ? 1 : fib(n - 1) + fib(n - 2);
}
```
parses to. Thus, there is no _speed_ advantage to defining the `fib` function
in the external fib.hpp file. However, it is possible to use an external C++
file to handle the gradient of a function analytically as opposed to using
Stan's autodifferentiation capabilities, which are slower and more memory
intensive but fully general. In this case, the `fib` function only deals with
integers so there is nothing to take the derivative of. The primary advantage
of using an external C++ file is flexibility to do things that cannot be done
directly in the Stan language. It is also useful for R packages like
__rstanarm__ that may want to define some C++ functions in the package's src
directory and rely on the linker to make them available in its Stan programs,
which are compiled at (or before) installation time.

In the C++ version, we check if `n` is non-positive, in which case we throw an
exception. The way the MCMC samplers are configured, if there is an exception
thrown and it is `std::domain_error`, it will treat that iteration as a rejection
but as a recoverable error: set that iteration's log probability value to negative
infinity and move to the next iteration. If there is a `std::invalid_argument`
thrown, MCMC terminates. We treat these as non-recoverable, user programming errors.
It is unnecessary to prefix `stringstream` with `std::` because of
the `using std::stringstream;` line in the _generated_ C++ file. However, there
is no corresponding `using std::domain_error;` line, so it has to be qualified
appropriately when the exception is thrown.

The only confusing part of the C++ version of the `fib` function is that it has
an additional argument (with no default value) named `pstream__` that is added
to the _declaration_ of the `fib` function by Stan. Thus, your _definition_ of
the `fib` function needs to match with this signature. This additional argument
is a pointer to a `std::ostream` and is only used if your function prints
something to the screen, which is rare. Thus, when we call the `fib` function
recursively in the last line, we specify `fib(n - 1, 0) + fib(n - 2, 0);` so
that the output (if any, and in this case there is none) is directed to the null
pointer.

This vignette has employed a toy example with the Fibonacci function, which has
little apparent use in a Stan program and if it were useful, would more easily
be implemented as a user-defined function in the `functions` block as
illustrated at the outset. The ability to use external C++ code only becomes
useful with more complicated C++ functions. It goes without saying that this
mechanism ordinarily cannot call functions in C, Fortran, R, or other languages
because Stan needs the derivatives with respect to unknown parameters in order
to perform estimation. These derivatives are handled with custom C++ types that
cannot be processed by functions in other languages that only handle primitive
types such as `double`, `float`, etc.

That said, it is possible to accomplish a great deal in C++, particularly when
utilizing the Stan Math Library. For more details, see
[The Stan Math Library: Reverse-Mode Automatic Differentiation in C++](https://arxiv.org/abs/1509.07164)
and its GitHub [repository](https://github.com/stan-dev/math/). The functions
that you _declare_ in the `functions` block of a Stan program will typically
involve templating and type promotion in their signatures when parsed to C++
(the only exceptions are functions whose only arguments are integers, as in
the `fib` function above). Suppose you wanted to define a function whose
arguments are real numbers (or at least one of the arguments is). For example,
```{r, eval = TRUE}
mc <-
'
functions { real sinc(real x); }
transformed data { real sinc_pi = sinc(pi()); }
'
stan_model(model_code = mc, model_name = "external", allow_undefined = TRUE,
           includes = paste0('\n#include "',
                             file.path(getwd(), 'sinc.hpp'), '"\n'))
```
The sinc.hpp file reads as
```{r, echo = FALSE, eval = TRUE, comment=""}
cat(readLines("sinc.hpp"), sep = "\n")
```
The body of the first `sinc` function is simply its mathematical definition in the
form of a ternary operator, which is sufficient when the input is a `double`.

The last lines of sinc.hpp are a specialization for when the input is an
unknown real parameter, which is represented in the Stan Math Library as a
`stan::math::var`. Since the derivative of the `sinc` function is easy to
compute analytically, we extract the underlying double-precision value from
the inputted `stan::math::var` and use that to calculate the function value
and its first derivative. To do this, we return the result of
`make_callback_vari()`, whose arguments are the function value (`sinc(x.val())`),
and a functor to update the derivative of `x` during the reverse-pass of the
autodifferentiation.

An easy way to see what the generated function signature will be is to call the
`stanc` function in the __rstan__ package with `allow_undefined = TRUE` and
inspect the resulting C++ code to see what function signatures needed to be
written for sinc.hpp.

Once you go to the trouble of writing a numerically stable C++ function, we
would welcome a pull request on GitHub to include your C++ function in the
Stan Math Library for everyone to benefit from, provided that it can be licensed
under the 3-clause BSD license and its use is not overly-specific to your
particular Stan program.

The Stan Math Library is compliant with the C++14 standard but not all compilers
fully support the C++14 standard. In particular, the compiler that comes with
RTools does not support all C++14 features. But you can use many C++14 features,
such as the `auto` keyword.
