---
title: "Simulation Based Calibration"
author: "Stan Development Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation Based Calibration}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
# Nothing
```

Here is a Stan program for a beta-binomial model
```{stan output.var="beta_binomial"}
data {
  int<lower = 1> N;
  real<lower = 0> a;
  real<lower = 0> b;
}
transformed data { // these adhere to the conventions above
  real pi_ = beta_rng(a, b);
  int y = binomial_rng(N, pi_);
}
parameters {
  real<lower = 0, upper = 1> pi;
}
model {
  target += beta_lpdf(pi | a, b);
  target += binomial_lpmf(y | N, pi);
}
generated quantities { // these adhere to the conventions above
  int y_ = y;
  vector[1] pars_;
  int ranks_[1] = {pi > pi_};
  vector[N] log_lik;
  pars_[1] = pi_;
  for (n in 1:y) log_lik[n] = bernoulli_lpmf(1 | pi);
  for (n in (y + 1):N) log_lik[n] = bernoulli_lpmf(0 | pi);
}
```
Notice that it adheres to the following conventions:
* Realizations of the unknown parameters are drawn in the `transformed data`
  block are postfixed with an underscore, such as `pi_`.
  These are considered the "true" parameters being estimated by
  the corresponding symbol declared in the `parameters` block, which
  have the same names except for the trailing underscore, such as `pi`.
* The realizations of the unknown parameters are then conditioned on when drawing from
  the prior predictive distribution in `transformed data` block, which in this
  case is `int y = binomial_rng(N, pi_);`. To avoid confusion, `y` does not have a 
  training underscore.
* The realizations of the unknown parameters are copied into a `vector`
  in the `generated quantities` block named `pars_`
* The realizations from the prior predictive distribution are copied 
  into an object (of the same type) in the `generated quantities` block
  named `y_. This is optional.
* The `generated quantities` block contains an integer array named 
  `ranks_` whose only values are zero or one, depending on whether the realization of a 
  parameter from the posterior distribution exceeds the corresponding "true"
  realization, which in this case is `ranks_[1] = {pi > pi_};`. These are not actually "ranks"
  but can be used afterwards to reconstruct (thinned) ranks.
* The `generated quantities` block contains a vector named `log_lik` whose values
  are the contribution to the log-likelihood by each observation. In this case, 
  the "observations" are the implicit successes and failures that are aggregated
  into a binomial likelihood. This is optional but facilitates calculating the
  Pareto k shape parameter estimates that indicate whether the posterior distribution
  is sensitive to particular observations.

Assuming the above is compile to a code `stanmodel` named `beta_binomial`, we can
then call the `SBC` function
```{r, eval = FALSE}
output <- SBC(beta_binomial, data = list(N = 10, a = 1, b = 1), M = 500, refresh = 0)
```
```{r, include = FALSE}
# This fakes what would happen if we actually took the time to run Stan.
N <- 10
M <- 500
pars_ <- rbeta(M, 1, 1)
y_ <- matrix(rbinom(pars_, size = N, prob = pars_), ncol = M)
post_ <- matrix(rbeta(M * 1000L, 1 + y_, 1 + N - y_), ncol = M)
ranks_ <- lapply(1:M, FUN = function(m) {
  matrix(post_[ , m] > pars_[m], ncol = 1, 
         dimnames = list(NULL, "pi"))
})
log_lik <- t(sapply(1:M, FUN = function(m) {
  c(dbinom(rep(1, y_[m]), size = 1, prob = pars_[m], log = TRUE),
    dbinom(rep(1, N - y_[m]), size = 1, prob = pars_[m], log = TRUE))
}))
sampler_params <- array(0, dim = c(1000, 6, M))
colnames(sampler_params) <- c("accept_stat__", "stepsize__", "treedepth__",   
                              "n_leapfrog__",  "divergent__",   "energy__")
output <- list(ranks = ranks_, Y = y_, pars = pars_, 
               log_lik = log_lik, sampler_params = sampler_params)
class(output) <- "SBC"
```
At which point, we can then call
```{r}
print(output)
plot(output, bins = 10) # it is best to specify the bins argument yourself
```

# References

  Talts, S., Betancourt, M., Simpson, D., Vehtari, A., and Gelman, A. (2018).
  Validating Bayesian Inference Algorithms with Simulation-Based Calibration.
  [arXiv preprint arXiv:1804.06788](https://arxiv.org/abs/1804.06788)
