---
title: "Accessing the contents of a stanfit object"
author: "Stan Development Team"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{stanfit-objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(rstan)
knitr::opts_chunk$set(echo = TRUE, comment = NA)
```

This vignette demonstrates how to access most of data stored in a stanfit 
object. A stanfit object (an object of class "stanfit") contains the output 
derived from fitting a Stan model using Markov chain Monte Carlo or one of
Stan's variational approximations (meanfield or full-rank). Throughout the
document we'll use the stanfit object obtained from fitting the Eight Schools
example model:

```{r, fit, results="hide", echo=FALSE}
J <- 8
y <- c(28,  8, -3,  7, -1,  1, 18, 12)
sigma <- c(15, 10, 16, 11,  9, 11, 10, 18)
fit <- stan(file="schools.stan",
             data = c("y", "J", "sigma"),
             iter=2000, chains=4, cores=1)
```
```{r, example-model, eval=FALSE}
fit <- stan_demo("eight_schools")
```

## Posterior draws

There are several functions that can be used to access the draws from the 
posterior distribution stored in a stanfit object. These are `extract`, 
`as.matrix`, `as.data.frame`, and `as.array`, each of which returns the draws in
a different format.

<br>

#### extract()

The `extract` function (with its default arguments) function returns a list with 
named components corresponding to the model parameters.

```{r, extract-1}
list_of_draws <- extract(fit)
print(names(list_of_draws))
```

In this model the parameters `mu` and `tau` are scalars and `theta` is a vector
with eight elements. This means that the draws for `mu` and `tau` will be 
vectors (with length equal to the number of post-warmup iterations times the 
number of chains) and the draws for `theta` will be a matrix, with each column 
corresponding to one of the eight components:

```{r, extract-2}
head(list_of_draws$mu)
head(list_of_draws$tau)
head(list_of_draws$theta)
```

<br>

#### as.matrix(), as.data.frame(), as.array()

The `as.matrix`, `as.data.frame`, and `as.array` functions can also be used
to retrieve the posterior draws from a stanfit object:

```{r, as.matrix-1}
matrix_of_draws <- as.matrix(fit)
print(colnames(matrix_of_draws))

df_of_draws <- as.data.frame(fit)
print(colnames(df_of_draws))

array_of_draws <- as.array(fit)
print(dimnames(array_of_draws))
```

The `as.matrix` and `as.data.frame` methods essentially return the same 
thing except in matrix and data frame form, respectively. The `as.array`
method returns the draws from each chain separately and so has an additional
dimension:

```{r, as.matrix-2, results="hold"}
print(dim(matrix_of_draws))
print(dim(df_of_draws))
print(dim(array_of_draws))
```

By default all of the functions for retrieving the posterior draws return the
draws for _all_ parameters (and generated quantities). The optional argument 
`pars` (a character vector) can be used if only a subset of the parameters is
desired, for example:

```{r, as.matrix-3}
mu_and_theta1 <- as.matrix(fit, pars = c("mu", "theta[1]"))
head(mu_and_theta1)
```

<br>

## Posterior summary statistics and convergence diagnostics

Summary statistics are obtained using the `summary` function. The object 
returned is a list with two components:

```{r, summary-1}
fit_summary <- summary(fit)
print(names(fit_summary))
```

In `fit_summary$summary` all chains are merged whereas `fit_summary$c_summary`
contains summaries for each chain individually. Typically we want the 
summary for all chains merged, which is what we'll focus on here.

The summary is a matrix with rows corresponding to parameters and columns to the
various summary quantities. These include the posterior mean, the posterior
standard deviation, and various quantiles computed from the draws. The `probs`
argument can be used to specify which quantiles to compute and `pars` can be
used to specify a subset of parameters to include in the summary.

For models fit using MCMC, also included in the summary are the Monte Carlo
standard error (`se_mean`), the effective sample size (`n_eff`), and the R-hat
statistic (`Rhat`).

```{r, summary-2}
print(fit_summary$summary)
```

If, for example, we wanted the only quantiles included to be 10% and 90%, and  for only the parameters included to be `mu` and `tau`, we would specify that like this:

```{r, summary-3}
mu_tau_summary <- summary(fit, pars = c("mu", "tau"), probs = c(0.1, 0.9))$summary
print(mu_tau_summary)
```

Since `mu_tau_summary` is a matrix we can pull out columns using their names:

```{r, summary-4}
mu_tau_80pct <- mu_tau_summary[, c("10%", "90%")]
print(mu_tau_80pct)
```


<br>

## Sampler diagnostics

For models fit using MCMC the  stanfit object will also contain the values of
parameters used for the sampler. The `get_sampler_params` function can 
be used to access this information.

The object returned by `get_sampler_params` is a list with one component (a
matrix) per chain. Each of the matrices has number of columns corresponding to
the number of sampler parameters and the column names provide the parameter
names. The optional argument inc_warmup (defaulting to `TRUE`) indicates whether
to include the warmup period.

```{r, get_sampler_params-1}
sampler_params <- get_sampler_params(fit, inc_warmup = FALSE)
sampler_params_chain1 <- sampler_params[[1]]
colnames(sampler_params_chain1)
```

To do things like calculate the average value of `accept_stat__` for each chain 
(or the maximum value of `treedepth__` for each chain if using the NUTS
algorithm, etc.) the `sapply` function is useful as it will apply the 
same function to each component of `sampler_params`:

```{r, get_sampler_params-2}
mean_accept_stat_by_chain <- sapply(sampler_params, function(x) mean(x[, "accept_stat__"]))
print(mean_accept_stat_by_chain)

max_treedepth_by_chain <- sapply(sampler_params, function(x) max(x[, "treedepth__"]))
print(max_treedepth_by_chain)
```

<br>

## Model code

The Stan program itself is also stored in the stanfit object and can be 
accessed using `get_stancode`:

```{r, get_stan_code-1}
code <- get_stancode(fit)
```

The object `code` is a single string and is not very intelligible when printed:

```{r, get_stan_code-2}
print(code)
```

A readable version can be printed using `cat`:
```{r, get_stan_code-3}
cat(code)
```

<br>

## Initial values

The `get_inits` function returns initial values as a list with one component per
chain. Each component is itself a (named) list containing the initial values for each parameter for the corresponding chain:

```{r, get_inits}
inits <- get_inits(fit)
inits_chain1 <- inits[[1]]
print(inits_chain1)
```

<br>

## (P)RNG seed

The `get_seed` function returns the (P)RNG seed as an integer:

```{r, get_seed}
print(get_seed(fit))
```

<br>

## Warmup and sampling times

The `get_elapsed_time` function returns a matrix with the warmup and sampling
times for each chain: 

```{r, get_elapsed_time}
print(get_elapsed_time(fit))
```
